```{r}
# =========================
# Libraries & helpers
# =========================
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(forecast)
  library(tseries)
  library(ggplot2)
  library(vars)
  library(urca)
})

safe_accuracy <- function(fc, ytest) {
  A <- accuracy(fc, ytest)
  A[, intersect(c("ME","RMSE","MAE","MPE","MAPE","MASE","ACF1","Theil's U"),
                colnames(A)), drop = FALSE]
}

adf_k <- function(x) trunc((length(na.omit(x)) - 1)^(1/3))

```
2. Load data & build ts objects
```{r}
# =========================
# Load & clean data
# =========================
output_directory <- "C:/nottingham 25-26/Math Group Project/CW2 - AGRICULTURE"
file_name <- "original.csv"
full_input_path <- paste0(output_directory, "/", file_name)

df <- read_csv(full_input_path) 
# , stringsAsFactors = FALSE

norm_str <- function(x) tolower(trimws(as.character(x)))
df <- df %>%
  mutate(
    series = norm_str(series),
    sector = norm_str(sector),
    value  = as.numeric(gsub(",", "", as.character(value)))
  )

d0 <- suppressWarnings(dmy(df$date))
d1 <- suppressWarnings(ymd(df$date))
d2 <- suppressWarnings(mdy(df$date))
df$date <- coalesce(d0, d1, d2)
stopifnot(!all(is.na(df$date)))

abs_df <- df %>%
  filter(series == "abs", !is.na(value)) %>%
  arrange(date)

# =========================
# Build quarterly ts for p0, p1, share
# =========================
gdp  <- abs_df %>%
  dplyr::filter(sector %in% c("p0","gdp","gross domestic product","jumlah","kdnk")) %>%
  dplyr::arrange(date) %>%
  dplyr::select(date, p0 = value)

agri <- abs_df %>%
  dplyr::filter(sector %in% c("p1","agriculture","agriculture, forestry & fishing","pertanian")) %>%
  dplyr::arrange(date) %>%
  dplyr::select(date, p1 = value)
dat <- inner_join(gdp, agri, by = "date") %>%
  arrange(date)

ts_p0 <- ts(dat$p0,
            start = c(year(min(dat$date)), quarter(min(dat$date))),
            frequency = 4)
ts_p1 <- ts(dat$p1, start = start(ts_p0), frequency = 4)
ts_share <- ts(100 * dat$p1 / dat$p0,
               start = start(ts_p0), frequency = 4)

tq <- time(ts_p0)

```
3. Trend + seasonality + pulses (same for VAR and Johansen)
```{r}
# =========================
# Deterministic terms (trend, season, pulses)
# =========================

# linear time trend
trend_full <- 1:length(ts_p0)

# quarterly seasonal dummies (3 cols)
season_full <- seasonaldummy(ts_p0)   # uses ts frequency = 4

# pulses
X_pulses <- cbind(
  covid_q2     = as.numeric(tq == 2020 + 1/4),
  covid_q3     = as.numeric(tq == 2020 + 2/4),
  shock_2022q2 = as.numeric(tq == 2022 + 1/4)
)

# combine all exogenous deterministics
exog_full <- cbind(trend = trend_full,
                   season_full,
                   X_pulses)

# logs for cointegration
log_p0 <- log(ts_p0)
log_p1 <- log(ts_p1)

y_levels <- cbind(log_p0, log_p1)
colnames(y_levels) <- c("log_p0", "log_p1")

```
4. Train / test split (same cut you’ve been using)
```{r}
# =========================
# Train / test split
# Train: up to 2023Q4
# Test : 2024Q1–2025Q4
# =========================
cut_year <- c(2023, 4)
cut_time <- 2023 + 3/4

y_train <- window(y_levels, end   = cut_year)
y_test  <- window(y_levels, start = c(2024, 1))

train_mask <- (time(ts_p0) <= cut_time)
test_mask  <- !train_mask

exog_train <- exog_full[train_mask, , drop = FALSE]
exog_test  <- exog_full[test_mask,  , drop = FALSE]

stopifnot(nrow(y_train) == nrow(exog_train))

```

5. Choose lag order on levels (with trend+season+pulses)
```{r}
# =========================
# VAR lag selection on TRAIN (levels)
# =========================
lag_sel <- VARselect(y_train,
                     lag.max = 4,
                     type    = "const",      # intercept in each eq.
                     exogen  = exog_train)   # trend + season + pulses

print(lag_sel$selection)
p_opt <- lag_sel$selection["AIC(n)"]   # or "SC(n)" for BIC choice

```
6. Johansen cointegration test (decide VAR vs VECM)
```{r}
# =========================
# Johansen cointegration test (TRAIN, levels)
# =========================
jo_tr <- ca.jo(y_train,
               type   = "trace",
               ecdet  = "const",    # constant in cointegration relation
               K      = 2,          # outputn for lagmax 4 =1 
               dumvar = exog_train) # SAME deterministics as above

summary(jo_tr)


# Quick decision at 5% level
jo_sum <- summary(jo_tr)
teststat <- jo_sum@teststat
critvals <- jo_sum@cval[,"5pct"]

stat_r0   <- teststat["r = 0"]
crit_r0   <- critvals["r = 0"]
stat_rle1 <- teststat["r <= 1"]
crit_rle1 <- critvals["r <= 1"]



```

trend test
```{r}

t <- 1:length(log_p0)

summary(lm(log_p0 ~ t))
summary(lm(log_p1 ~ t))

```


Less XReg
```{r}
# Logs: levels
adf_logp0_lvl    <- adf.test(log_p0,    k = adf_k(log_p0))
adf_logp1_lvl    <- adf.test(log_p1,    k = adf_k(log_p1))

adf_logp0_lvl
adf_logp1_lvl

S_train  <- season_full[train_mask, , drop = FALSE]


p_opt <- VARselect(y_train, lag.max = 4, type="trend")$selection["AIC(n)"]
summary(p_opt)
K_joh <- max(2, p_opt + 1)

S_train  <- season_full[train_mask, , drop = FALSE]

trend_joh <- ca.jo(
  y_train,
  type   = "trace",
  ecdet  = "trend",  # constant in cointegration space
  K      = K_joh,
  dumvar = S_train   # ONLY seasonal dummies
)

summary(trend_joh)


```
Johansen Result:
Trace test suggests 1 cointegration vector (r = 1).

However, this test is known to be unreliable in small samples (T ≈ 40) and
in the presence of strong deterministic trends and structural breaks (e.g.,
COVID-19). Therefore, the long-run vector must be validated using the
Engle–Granger residual test.
Cointegration Residual Test
```{r}

trend_vec <- 1:nrow(y_train)
trend_ts <- ts(1:nrow(y_train), start=start(y_train), frequency=4)
Y_aug <- cbind(
  log_p0 = y_train[,1],
  log_p1 = y_train[,2],
  trend  = trend_ts
)

beta <- trend_joh@V[,1]    # first cointegration vector
ect  <- Y_aug %*% beta      # NOW dimensions match
ect  <- ts(ect, start=start(y_train), freq=4)


autoplot(ect) +
  ggtitle("Cointegration Residual (EC term)") +
  theme_minimal()

adf.test(ect)

t_res <- 1:length(ect)
summary(lm(as.numeric(ect) ~ t_res))
```


## 4.0 Testing for Cointegration Between GDP (p0) and Agriculture GDP (p1)

This section evaluates whether total GDP (p0) and agriculture GDP (p1) share a
long-run equilibrium relationship. If the two series are cointegrated, a Vector
Error Correction Model (VECM) is required. If not, a standard Vector
Autoregression (VAR) in differences is the appropriate model.

We use:
- **Logs of p0 and p1**
- **Seasonal dummies**
- **Train sample only: 2015Q1–2023Q4**

Intervention dummies (COVID-19 and 2022 shocks) are *not included* in the
cointegration test because the Johansen procedure assumes no large structural
breaks.

---

### 4.1 Step 1 — Stationarity Checks (ADF)

```
library(tseries)

adf_logp0 <- adf.test(log_p0)
adf_logp1 <- adf.test(log_p1)

adf_logp0_sdiff <- adf.test(diff(log_p0, lag = 4))
adf_logp1_sdiff <- adf.test(diff(log_p1, lag = 4))

adf_logp0
adf_logp1
adf_logp0_sdiff
adf_logp1_sdiff
```


❌ Your cointegration vector IS NOT stationary

→ NO COINTEGRATION
→ You should NOT use a VECM
→ Use VAR (in seasonal differences) instead.

✅ 1. ADF Test on ECT (MUST be stationary)
ADF = -1.7168  
p-value = 0.6833  

That means:

❌ Fail to reject H₀ (nonstationary)

The EC term has a unit root.
📌 If ECT is nonstationary → NO valid cointegration relationship → VECM is invalid.



✅ 2. Trend regression on ECT

Results:

Trend coefficient p = 0.28 → not significant
BUT the residual still has variance = 0.072, quite large
No visible mean-reversion pattern

This implies:
No trending behavior (good)
But ECT does not revert to a long-run equilibrium (bad)

A valid cointegration residual should oscillate tightly around 0, not drift around 14.8–15.1 with huge COVID shock dips.


✅ 3. Visual inspection (your plot)

Your EC-term:

Looks nonstationary
Shows persistent shifts
COVID shock creates random walk-type behavior
No strong pullback to any equilibrium line


4.5 Why Johansen Gives a False Positive

Although Johansen indicated r = 1, this is a known problem in small samples:

Both series contain strong linear trends

Sample size is very small (n ≈ 40)

There are large structural breaks (COVID-19)

Johansen is highly sensitive to lag selection

These issues inflate the trace statistic, causing false detection of
cointegration.

The Engle–Granger test correctly identifies no cointegration.



