---
title: "p2finalmaybe"
author: "Leong Jun Chuen"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

```{r}
# ---- Libraries ----
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(forecast)
  library(tseries)
  library(ggplot2)
})

# ---- Helpers ----
safe_accuracy <- function(fc, ytest) {
  A <- accuracy(fc, ytest)
  A[, intersect(c("ME","RMSE","MAE","MPE","MAPE","MASE","ACF1","Theil's U"),
                colnames(A)), drop = FALSE]
}
lb_p <- function(fit, y, lag = 8) {
  k <- length(coef(fit)); if (is.null(k)) k <- 0
  Box.test(residuals(fit), type = "Ljung-Box",
           lag = min(lag, length(y)-1), fitdf = k)$p.value
}
get_AICc <- function(fit) {  # robust AICc
  if (!is.null(fit$aicc)) return(fit$aicc)
  aic <- AIC(fit); n <- length(residuals(fit)); k <- length(coef(fit))
  if (is.finite(n) && is.finite(k) && (n > k + 1))
    aic + (2 * k * (k + 1)) / (n - k - 1) else aic
}
plot_backtest <- function(ts_full, fc, start_yq, title) {
  n  <- length(ts_full)
  f  <- frequency(ts_full)
  st <- start(ts_full)
  all_t <- st[1] + (0:(n-1) + st[2]-1)/f
  # quick quarterly date index
  q_to_month <- function(q) c(1,4,7,10)[q]
  yy <- floor(all_t)
  qq <- (round((all_t - yy)*f) %% f) + 1
  dates <- as.Date(sprintf("%04d-%02d-01", yy, q_to_month(qq)))
  
  test <- window(ts_full, start = start_yq)
  idx  <- (n - length(test) + 1):n
  
  df <- tibble(
    date  = dates[idx],
    actual= as.numeric(test),
    mean  = as.numeric(fc$mean),
    lo80  = as.numeric(fc$lower[,"80%"]), hi80 = as.numeric(fc$upper[,"80%"]),
    lo95  = as.numeric(fc$lower[,"95%"]), hi95 = as.numeric(fc$upper[,"95%"])
  )
  
  ggplot(df, aes(date, actual)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = .12) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), alpha = .20) +
    geom_line(size = 0.8) +
    geom_line(aes(y = mean), linewidth = 0.9, linetype = "dashed") +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 12)
}

adf_k <- function(x) trunc((length(na.omit(x)) - 1)^(1/3))


# ---- Load & clean ----
df <- read.csv("DAta.csv", stringsAsFactors = FALSE)

norm_str <- function(x) tolower(trimws(as.character(x)))
df <- df %>%
  mutate(
    series = norm_str(series),
    sector = norm_str(sector),
    value  = as.numeric(gsub(",", "", as.character(value)))
  )

d0 <- suppressWarnings(dmy(df$date))
d1 <- suppressWarnings(ymd(df$date))
d2 <- suppressWarnings(mdy(df$date))
df$date <- coalesce(d0, d1, d2)
stopifnot(!all(is.na(df$date)))

abs_df <- df %>% filter(series == "abs", !is.na(value)) %>% arrange(date)

# ---- Build p0, p1, share (quarterly) ----
gdp  <- abs_df %>% filter(sector %in% c("p0","gdp","gross domestic product","jumlah","kdnk")) %>%
  arrange(date) %>% select(date, p0 = value)
agri <- abs_df %>% filter(sector %in% c("p1","agriculture","agriculture, forestry & fishing","pertanian")) %>%
  arrange(date) %>% select(date, p1 = value)

dat <- inner_join(gdp, agri, by = "date") %>% arrange(date)

ts_p0 <- ts(dat$p0, start = c(year(min(dat$date)), quarter(min(dat$date))), frequency = 4)
ts_p1 <- ts(dat$p1, start = start(ts_p0), frequency = 4)
ts_share <- ts(100 * dat$p1/dat$p0, start = start(ts_p0), frequency = 4)

# ==== Plots for p0, p1, and share ====

# Tidy frame for plotting
plot_dat <- dat %>%
  mutate(share = 100 * p1/p0) %>%
  rename(`GDP (p0)` = p0,
         `Agriculture (p1)` = p1,
         `Share of GDP (%)` = share)

# 1) Individual charts -------------------------------------------------------
p_p0 <- ggplot(plot_dat, aes(date, `GDP (p0)`)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Total GDP (p0)", x = NULL, y = NULL) +
  theme_minimal(base_size = 12)

p_p1 <- ggplot(plot_dat, aes(date, `Agriculture (p1)`)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Agriculture GDP (p1)", x = NULL, y = NULL) +
  theme_minimal(base_size = 12)

p_share <- ggplot(plot_dat, aes(date, `Share of GDP (%)`)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Agriculture share of GDP (%)", x = NULL, y = NULL) +
  theme_minimal(base_size = 12)

# Print to the plot pane
#p_p0; p_p1; p_share

# 2) Faceted summary (each series on its own y-axis) -------------------------
plot_long <- plot_dat |>
  pivot_longer(cols = c(`GDP (p0)`, `Agriculture (p1)`, `Share of GDP (%)`),
               names_to = "series", values_to = "value")

p_facets <- ggplot(plot_long, aes(date, value)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(title = "GDP, Agriculture, and Share over time", x = NULL, y = NULL) +
  theme_minimal(base_size = 12)

p_facets

# (Optional) mark COVID quarters (2020Q2, 2020Q3)
covid_lines <- as.Date(c("2020-04-01", "2020-07-01"))
p_p1 + geom_vline(xintercept = covid_lines, linetype = "dashed", alpha = .4)
p_share + geom_vline(xintercept = covid_lines, linetype = "dashed", alpha = .4)
p_p0 + geom_vline(xintercept = covid_lines, linetype = "dashed", alpha = .4)


tq <- time(ts_p1)
X <- cbind(
  covid_q2     = as.numeric(tq == 2020 + 1/4),
  covid_q3     = as.numeric(tq == 2020 + 2/4),
  shock_2022q2 = as.numeric(tq == 2022 + 1/4)
)
# (Optional) Save to files
# ggsave("plot_p0.png", p_p0, width = 8, height = 4, dpi = 150)
# ggsave("plot_p1.png", p_p1, width = 8, height = 4, dpi = 150)
# ggsave("plot_share.png", p_share, width = 8, height = 4, dpi = 150)
# ggsave("plot_facets.png", p_facets, width = 9, height = 7, dpi = 150)


# =========================
# Pulses / Intervention Dummies (xreg)
# =========================
tq <- time(ts_p0)

X <- cbind(
  covid_q2     = as.numeric(tq == 2020 + 1/4),  # 2020Q2
  covid_q3     = as.numeric(tq == 2020 + 2/4),  # 2020Q3
  shock_2022q2 = as.numeric(tq == 2022 + 1/4)   # 2022Q2
)

X_ts <- ts(X, start = start(ts_p0), frequency = 4)

# ---- A quick look ----
ndiffs(ts_share); nsdiffs(ts_share)
ggAcf(ts_share); ggPacf(ts_share)   # uncomment to view
```

```{r}

# Put the two log series into a tibble
df_plot <- tibble(
  date = as.Date(time(log_p0)),
  log_p0 = as.numeric(log_p0),
  log_p1 = as.numeric(log_p1)
) %>%
  pivot_longer(cols = c(log_p0, log_p1),
               names_to = "series",
               values_to = "value")

# Plot
ggplot(df_plot, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 1) +
  labs(title = "Log(p0) vs Log(p1)",
       x = NULL,
       y = "Log value",
       color = "Series") +
  theme_minimal(base_size = 12)
```


## Transformation

```{r}
# ===== Apply transformations =====
# Log transformations
log_p1 <- log(ts_p1)
log_p0 <- log(ts_p0)

# Share with log-transformed variables
log_share <- log(ts_p1) / log(ts_p0)  # log(p1/p0)
log_share_tgt <- log(ts_p1) - log(ts_p0)  # log(p1/p0)

# ---- Quick Check of Transformations ----
ndiffs(log_share); nsdiffs(log_share)   # Expect d=0, D=1 for log share
autoplot(log_share) + ggtitle("Log Agriculture Share of GDP (%)")
autoplot(log_p1) + ggtitle("Log Agriculture ")
autoplot(log_p0) + ggtitle("Log GDP")

# ---- ACF/PACF ----
ggAcf(log_share)
ggPacf(log_share)

# ADF Test with differencing and lag 4 for log-transformed series
adf_logp1_sdiff <- adf.test(diff(log(ts_p1), lag = 4), k = adf_k(log(ts_p1)))  # ADF on diff(log(p1))
adf_logp0_sdiff <- adf.test(diff(log(ts_p0), lag = 4), k = adf_k(log(ts_p0)))  # ADF on diff(log(p0))
adf_logshare_sdiff <- adf.test(diff(log(ts_share), lag = 4), k = adf_k(log(ts_share)))  # ADF on diff(log(share))

# ADF Test for non-transformed series (without differencing)
adf_p1_sdiff <- adf.test(diff(ts_p1, lag = 4), k = adf_k(ts_p1))  # ADF on diff(p1)
adf_p0_sdiff <- adf.test(diff(ts_p0, lag = 4), k = adf_k(ts_p0))  # ADF on diff(p0)
adf_share_sdiff <- adf.test(diff(ts_share, lag = 4), k = adf_k(ts_share))  # ADF on diff(share)

# ADF Test without differencing for log-transformed series
adf_logp1_lvl <- adf.test(log(ts_p1), k = adf_k(log(ts_p1)))  # ADF on log(p1) without differencing
adf_logp0_lvl <- adf.test(log(ts_p0), k = adf_k(log(ts_p0)))  # ADF on log(p0) without differencing
adf_logshare_lvl <- adf.test(log(ts_share), k = adf_k(log(ts_share)))  # ADF on log(share) without differencing

# ADF Test for non-transformed data (without differencing)
adf_p1_lvl <- adf.test(ts_p1, k = adf_k(ts_p1))  # ADF on p1 without differencing
adf_p0_lvl <- adf.test(ts_p0, k = adf_k(ts_p0))  # ADF on p0 without differencing
adf_share_lvl <- adf.test(ts_share, k = adf_k(ts_share))  # ADF on share without differencing

# Print results for both log and non-log series
print(adf_logp1_sdiff)  # Log(p1) with differencing
print(adf_logp0_sdiff)  # Log(p0) with differencing
print(adf_logshare_sdiff)  # Log(share) with differencing

print(adf_logp1_lvl)  # Log(p1) without differencing
print(adf_logp0_lvl)  # Log(p0) without differencing
print(adf_logshare_lvl)  # Log(share) without differencing

# Print results for non-log series
print(adf_p1_sdiff)  # p1 with differencing
print(adf_p0_sdiff)  # p0 with differencing
print(adf_share_sdiff)  # share with differencing

print(adf_p1_lvl)  # p1 without differencing
print(adf_p0_lvl)  # p0 without differencing
print(adf_share_lvl)  # share without differencing

```

## Stationarity and Cointegration Diagnostics

In this section, we assess the time series properties of the GDP series:

-   $p_0$: total GDP\
-   $p_1$: agriculture GDP\
-   $\text{share} = 100 \times p_1/p_0$
-   and their log–transformations $\log(p_0)$, $\log(p_1)$,
    $\log(\text{share})$.

We use Augmented Dickey–Fuller (ADF) tests to check for
(non)-stationarity and the Johansen test to check for cointegration
between $\log(p_0)$ and $\log(p_1)$.

```{r adf-setup, echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
library(ggplot2)

# Levels
adf_p0_lvl    <- adf.test(ts_p0,    k = adf_k(ts_p0))
adf_p1_lvl    <- adf.test(ts_p1,    k = adf_k(ts_p1))
adf_share_lvl <- adf.test(ts_share, k = adf_k(ts_share))   # you ran this separately

adf_p0_lvl
adf_p1_lvl
adf_share_lvl

# Seasonal differences (D = 1, quarterly)
adf_p0_sdiff    <- adf.test(diff(ts_p0,    lag = 4), k = adf_k(ts_p0))
adf_p1_sdiff    <- adf.test(diff(ts_p1,    lag = 4), k = adf_k(ts_p1))
adf_share_sdiff <- adf.test(diff(ts_share, lag = 4), k = adf_k(ts_share))

adf_p0_sdiff
adf_p1_sdiff
adf_share_sdiff


```

-   **Summary (non-log series, based on your output):**

-   **Share in levels `tsshare`** :ADF statistic ≈ −1.81, p-value ≈
    **0.65** → **fail to reject** the unit-root null.⇒ The share series
    is **non-stationary in levels**.

-   **Seasonally differenced series `diff(ts_p1, lag = 4)` and
    `diff(ts_share, lag = 4)`**:You obtained p-values around
    **0.02–0.09**, which are **much lower** than in levels and
    consistent with the seasonal differencing identified by nsdiffs()⇒
    After seasonal differencing (lag 4), the series behave much closer
    to **stationary**.

```{r}
# Logs: levels
adf_logp0_lvl    <- adf.test(log_p0,    k = adf_k(log_p0))
adf_logp1_lvl    <- adf.test(log_p1,    k = adf_k(log_p1))
adf_logshare_lvl <- adf.test(log_share, k = adf_k(log_share))

adf_logp0_lvl
adf_logp1_lvl
adf_logshare_lvl

# Logs: seasonal difference (lag = 4)
adf_logp0_sdiff    <- adf.test(diff(log_p0,    lag = 4), k = adf_k(log_p0))
adf_logp1_sdiff    <- adf.test(diff(log_p1,    lag = 4), k = adf_k(log_p1))
adf_logshare_sdiff <- adf.test(diff(log_share, lag = 4), k = adf_k(log_share))

adf_logp0_sdiff
adf_logp1_sdiff
adf_logshare_sdiff

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(knitr)

############################################
# 1) LOG SERIES (LEVELS)
############################################

log_levels <- data.frame(
  Series = c("log(p1)", "log(p0)", "log(share)"),
  ADF_Statistic = c(-1.8918, -1.8690, -1.7707),
  Lag = 3,
  p_value = c(0.6163, 0.6253, 0.6640),
  Stationary = c("No", "No", "No")
)

kable(log_levels,
      caption = "ADF Test: Log-Transformed Series (No Differencing)",
      digits = 5, align = "c")


############################################
# 2) LOG SERIES (SEASONALLY DIFFERENCED, LAG 4)
############################################

log_diff4 <- data.frame(
  Series = c("diff(log(p1))", "diff(log(p0))", "diff(log(share))"),
  ADF_Statistic = c(-3.9482, -3.2645, -3.4654),
  Lag = 3,
  p_value = c(0.02231, 0.09218, 0.06271),
  Stationary = c("Yes*", "Borderline", "Borderline")
)

kable(log_diff4,
      caption = "ADF Test: Log-Transformed Series with Seasonal Differencing (lag = 4)",
      digits = 5, align = "c")


############################################
# 3) RAW SERIES (LEVELS, NO LOG, NO DIFFERENCING)
############################################

raw_levels <- data.frame(
  Series = c("p1", "p0", "share"),
  ADF_Statistic = c(-1.9530, -1.7279, -1.8086),
  Lag = 3,
  p_value = c(0.5922, 0.6809, 0.6491),
  Stationary = c("No", "No", "No")
)

kable(raw_levels,
      caption = "ADF Test: Raw Series (No Differencing)",
      digits = 5, align = "c")


############################################
# 4) RAW SERIES (SEASONALLY DIFFERENCED, LAG 4)
############################################

raw_diff4 <- data.frame(
  Series = c("diff(p1)", "diff(p0)", "diff(share)"),
  ADF_Statistic = c(-4.0247, -3.3309, -3.5000),
  Lag = 3,
  p_value = c(0.01938, 0.08244, 0.05762),
  Stationary = c("Yes*", "Borderline", "Borderline")
)

kable(raw_diff4,
      caption = "ADF Test: Raw Series with Seasonal Differencing (lag = 4)",
      digits = 5, align = "c")

```




#### **Table 1 — Log levels (not stationary)**
All log-level series have very high p-values (>0.60) → clearly **non-stationary**.

#### **Table 2 — Log-differenced seasonal**
- log(p1) is **stationary** after seasonal difference  
- log(p0) and log(share) are **borderline** (close to 5%)

#### **Table 3 — Raw levels (not stationary)**
p1, p0, share all fail stationarity (p > 0.59).

#### **Table 4 — Raw seasonal diff**
- p1 becomes **stationary** (p = 0.019)  
- p0 and share are borderline

---

**Conclusion:**  
All three series require **seasonal log-differencing (Δ₄ log())** before modelling → which matches your VARMAX specification.


## 4. Stationarity, Cointegration, and Model Justification

Before specifying the multivariate time-series model, the level series
for total GDP ($p_0$) and agriculture GDP ($p_1$) were transformed into
logarithms. Unit root tests using the Augmented Dickey–Fuller (ADF)
procedure indicate that both $\log(p_0)$ and $\log(p_1)$ are
non–stationary in levels (ADF $p$–values $>$ 0.60).

Seasonal differencing of order four was applied to reflect the quarterly
structure of the data:

$$ \Delta_4 \log(p_t) = \log(p_t) - \log(p_{t-4}). $$

After this transformation, the ADF tests show substantially lower
$p$–values (approximately 0.02–0.09), indicating that the seasonal
log-differences are stationary or near–stationary. This implies the
series behave like I(1) processes with seasonal integration.

To evaluate whether the two GDP components share a long–run equilibrium,
the Johansen cointegration test was applied to the vector
$\left[\log(p_0),\ \log(p_1)\right]$. Both the trace and
maximum–eigenvalue statistics fail to reject the null hypothesis of no
cointegration ($r=0$). Therefore, there is no evidence of a stable
long–run cointegration relationship between the two logged GDP series.

The implications for modelling are clear:

-   Because the variables are non–stationary in levels but **not
    cointegrated**, a **VECM is not appropriate**.
-   Instead, the correct specification is a VAR model estimated on the
    **stationary transformed series** $\Delta_4\log(p_0)$ and
    $\Delta_4\log(p_1)$.

Moreover, the series contain identifiable structural shocks in 2020Q2–Q3
(COVID–19 lockdowns) and 2022Q2 (post-COVID agricultural spike). These
are better handled through exogenous intervention dummies rather than
forcing the VAR dynamics to absorb large outliers. Therefore, a VAR with
exogenous regressors (VARX) is the preferred specification.

```{r pressure, echo=FALSE}
library(urca)

y_pair <- cbind(log_p1, log_p0)

# Example specification matching your output:
jo_test <- ca.jo(y_pair,
                 type  = "trace",
                 K     = 2,
                 ecdet = "const")  # constant in cointegration space

summary(jo_test)
```

-   For **r = 0** (no cointegration):test = 14.71 \< 17.85 (10% critical
    value)⇒ **Fail to reject H₀ of no cointegration.**

-   For **r ≤ 1**:test = 4.68 \< 7.52 (10% critical value)⇒ Also fail to
    reject.

**Conclusion (cointegration):**

Under this specification, the Johansen trace test **does not find
evidence of cointegration** between $log(p_1)$ and $log(p_0)$. In other
words, although both series are non-stationary in levels, they **do not
share a statistically significant long-run equilibrium relationship** in
the sense of cointegration.

Therefore:

-   It is appropriate to model **each series separately** using
    ARIMA/SARIMA in (seasonal) differences; and

-   For the **contribution share** (either in levels or logs), it is
    reasonable to model it directly (e.g. SARIMA or SARIMAX), rather
    than imposing a VECM structure based on a cointegrating relationship
    between $log⁡(p0$) and $log⁡(p1$).

```{r}
library(vars)
library(lmtest)
# --- Log levels ---
log_p0 <- log(ts_p0)
log_p1 <- log(ts_p1)

# --- Seasonal differences (D = 1, quarterly) ---
d4_log_p0 <- diff(ts_p0, lag = 4)  # approx yearly log growth of total GDP
d4_log_p1 <- diff(ts_p1, lag = 4)  # approx yearly log growth of agriculture GDP

# Put into a bivariate ts object for VAR
start0 <- start(ts_p0)  # e.g. c(2010, 1)
y_var <- ts(
  cbind(d4_log_p0, d4_log_p1),
  start = c(start0[1] + 1, start0[2]),  # 1 year later because lag = 4
  frequency = 4
)
colnames(y_var) <- c("d4_log_p0", "d4_log_p1")

# Time index for differenced series
t_y <- time(y_var)

# Match rows of X to the times in y_var
X_var <- X[match(t_y, tq), , drop = FALSE]

```


## 4.0 Methodology: VARMAX Modelling of Agriculture GDP and Total GDP

### 4.1 Selection of Optimal Lag Order

```{r}
# Choose lag order (e.g. up to 4 lags for quarterly data)
lag_sel <- VARselect(y_var, lag.max = 4, type = "const", exogen = X_var)
lag_sel

p_opt <- lag_sel$selection["AIC(n)"]  # or choose BIC if you prefer


```



The optimal lag length was chosen using the `VARselect()` function in R, which evaluates several information criteria:

- Akaike Information Criterion (AIC)
- Hannan–Quinn Criterion (HQ)
- Schwarz Criterion (SC)
- Final Prediction Error (FPE)

`VARselect()` recommended:

- **lag = 4** according to AIC, HQ, and FPE  
- **lag = 1** according to SC  

Since AIC, HQ, and FPE penalise residual structure less harshly and prioritize predictive accuracy,  
a **VAR(4)** model was chosen.

This is similar to how `auto.arima()` selects ARIMA orders—  
VARselect is the multivariate equivalent for VAR models.

### 4.3 Incorporating Exogenous Shocks (VARX / VARMAX Structure)

```{r}
var_fit <- VAR(
  y_var,
  p      = p_opt,
  type   = "const",
  exogen = X_var
)

summary(var_fit)

```




The pandemic (2020Q2–Q3) and the 2022Q2 shock produced large residual spikes in earlier ARIMA models.  
To avoid contaminating the VAR residual structure, three intervention dummy variables were introduced:

- `covid_q2` = 1 in 2020Q2, 0 otherwise  
- `covid_q3` = 1 in 2020Q3, 0 otherwise  
- `shock_2022q2` = 1 in 2022Q2, 0 otherwise  

Thus the system becomes a **VARMAX model**:

\[
y_t = c + A_1 y_{t-1} + \cdots + A_4 y_{t-4} + B X_t + u_t
\]

where:

- $y_t$ = vector of endogenous variables  
- $A_i$ = autoregressive coefficient matrices  
- $X_t$ = exogenous shock dummy variables  
- $B$ = matrix of exogenous effects  
- $u_t$ = vector of white-noise errors  

```{r}
#run this if dw exogenus
var_fit_nox <- VAR(y_var, p = p_opt, type = "const")
summary(var_fit_nox)

```



### Results and Dynamic Contribution of Agriculture to GDP

We estimate a bivariate VAR model for the seasonal log–growth rates of
total and agricultural GDP. Let $d4\_log\_p0 = \Delta_4 \log(p_0)$
denote the quarterly seasonal difference of total real GDP and
$d4\_log\_p1 = \Delta_4 \log(p_1)$ the corresponding series for
agricultural GDP. The lag order $p=4$ is chosen using AIC, HQ and FPE
criteria.

Two specifications are considered:

1.  A baseline VAR(4) with only a constant term.
2.  A VAR(4) with exogenous intervention dummies (`covid_q2`,
    `covid_q3`, `shock_2022q2`) capturing the COVID-19 recession in
    2020Q2–Q3 and the post-COVID rebound in 2022Q2.

#### VAR with exogenous interventions

For the equation in $d4\_log\_p0$, the coefficients on its own lags at
horizons 2–4 are significant, while the lags of $d4\_log\_p1$ are not.
The dummy `covid_q2` enters with a large negative and highly significant
coefficient, indicating a sharp contraction in aggregate GDP growth
during 2020Q2, whereas `shock_2022q2` is significantly positive,
consistent with a strong rebound in 2022Q2. The adjusted $R^2$ of this
equation is approximately 0.79.

In the equation for $d4\_log\_p1$, the first lag of agricultural growth
$d4\_log\_p1.l1$ is positive and highly significant, indicating
persistence in agricultural growth. The first lag of aggregate GDP
growth $d4\_log\_p0.l1$ is significantly negative, suggesting that
unusually strong aggregate growth in one quarter is followed by
relatively weaker agricultural growth in the next. The 2022Q2 dummy has
a positive, borderline-significant effect, while the COVID dummies are
not significant in this equation. The adjusted $R^2$ is about 0.73.

When the intervention dummies are omitted (baseline VAR), the adjusted
$R^2$ for the $d4\_log\_p0$ equation falls from roughly 0.79 to 0.23 and
the residual variance roughly doubles. This confirms that the large
COVID-19 and 2022Q2 movements are better modelled as exogenous
interventions than forced into the autoregressive dynamics. We therefore
treat the VAR with exogenous dummies as the preferred VARX
specification.


# 5. Check “contribution” of $p_1$ to $p_0$

## (a) Granger causality

Does agriculture growth Granger-cause total GDP growth?

```{r}
# Does d4_log_p1 help predict d4_log_p0?
causality(var_fit, cause = "d4_log_p1")

# (Optional) reverse direction:
causality(var_fit, cause = "d4_log_p0")

#If the p-value in the first test is small (< 0.05), you can say
#“Agricultural GDP growth Granger-causes total GDP growth in the VAR model.
```

Two null hypotheses were tested:

1. **$p_1$ does NOT Granger-cause $p_0$**  
   - p = 0.9539 → fail to reject H0  
   - **Agriculture does not drive total GDP**

2. **$p_0$ does NOT Granger-cause $p_1$**  
   - p = 0.0620 → marginal evidence  
   - **Total GDP may influence agriculture**, but weakly

Instantaneous causality tests showed **no contemporaneous correlation** between the shocks in both equations.




Granger causality tests based on the baseline VAR indicate that
agricultural growth does not Granger-cause aggregate GDP growth: the
F-test for the null hypothesis “$d4\_log\_p1$ does not Granger-cause
$d4\_log\_p0$” has a p-value of about 0.95, so the null cannot be
rejected. In contrast, the null “$d4\_log\_p0$ does not Granger-cause
$d4\_log\_p1$” yields a p-value around 0.06, which provides weak (10%
level) evidence that aggregate GDP growth helps predict future
agricultural growth.

Instantaneous causality tests, which examine the covariance of the VAR
residuals conditional on all lags, show no evidence of contemporaneous
dependence between $d4\_log\_p0$ and $d4\_log\_p1$ (Chi-squared p-value
≈ 0.89). After controlling for lagged values, shocks to agriculture and
total GDP within the same quarter appear to be statistically
independent.

## (b) Impulse Response Functions (IRF)

```{r}
irf_p1_to_p0 <- irf(
  var_fit,
  impulse  = "d4_log_p1",
  response = "d4_log_p0",
  n.ahead  = 8,   # 8 quarters ahead
  boot     = TRUE
)

plot(irf_p1_to_p0)
```

Orthogonal impulse response functions (IRFs) from the preferred VARX
model show that a one–standard deviation shock to agricultural growth
($d4\_log\_p1$) has only a very small effect on aggregate GDP growth
($d4\_log\_p0$) at all horizons. The 95% bootstrap confidence intervals
always include zero, implying that the dynamic effect is not
statistically significant.

Forecast error variance decompositions (FEVDs) further confirm this
result. At all horizons, almost all of the forecast error variance of
$d4\_log\_p0$ is explained by its own shocks, with only a very small
share attributable to shocks in $d4\_log\_p1$. Similarly, the variance
of $d4\_log\_p1$ is dominated by its own innovations. Thus, the
contribution of agriculture-specific shocks to the short-run variability
of aggregate GDP appears to be limited.

## (c) Forecast Error Variance Decomposition (FEVD)

```{r}
fevd_fit <- fevd(var_fit, n.ahead = 8)
plot(fevd_fit)


```

Overall, the VAR analysis suggests that, in quarterly seasonal
log-growth terms, agricultural GDP does not exert a strong dynamic
influence on aggregate GDP in Malaysia over the sample period. The main
determinants of total GDP growth are its own past values and large
exogenous shocks such as the COVID-19 recession and subsequent rebound.
There is some weak evidence that aggregate GDP growth helps to predict
agricultural growth, but impulse responses and variance decompositions
indicate that these cross-effects are economically small.



## 5. Final Model Choice: VAR/VARX in Seasonal Log-Differences

The final multivariate specification is a VAR($p$) model for quarterly
GDP growth components:

$$
\begin{bmatrix}
\Delta_4 \log(p_{0,t}) \\
\Delta_4 \log(p_{1,t})
\end{bmatrix}
=
A_1
\begin{bmatrix}
\Delta_4 \log(p_{0,t-1}) \\
\Delta_4 \log(p_{1,t-1})
\end{bmatrix}
+ \cdots +
A_p
\begin{bmatrix}
\Delta_4 \log(p_{0,t-p}) \\
\Delta_4 \log(p_{1,t-p})
\end{bmatrix}
+ B X_t + \varepsilon_t,
$$

where $X_t$ contains intervention dummies for:

-   2020Q2 (COVID shock),
-   2020Q3 (COVID shock),
-   2022Q2 (post-pandemic agricultural spike).

Using standard information criteria (AIC, HQ and FPE), the optimal lag
length is $p = 4$, while the Schwarz Criterion suggests $p = 1$. Given
the quarterly economic structure and the agreement of three criteria,
the VAR(4) specification is adopted.

This VARX(4) model is equivalent to a VARMAX model with only
autoregressive terms, providing:

-   Granger causality analysis to test whether agriculture growth
    predicts GDP growth,
-   impulse response functions (IRFs) to quantify dynamic effects of
    shocks to $p_1$ on $p_0$,
-   forecast error variance decomposition (FEVD) to measure the
    contribution of agriculture fluctuations to GDP variability.

The absence of cointegration, combined with seasonal stationarity and
identifiable exogenous shocks, supports the VAR/VARX framework as the
most theoretically consistent and empirically suitable approach for
analysing the contribution of agriculture to GDP.

------------------------------------------------------------------------









### 4.4 Estimation Results

The VARMAX(4) estimate produces two equations:

- Equation 1: $\Delta_4 \log(p_{0,t})$  
- Equation 2: $\Delta_4 \log(p_{1,t})$

Key findings:

#### (a) Total GDP equation ($\Delta_4 \log(p_0)$)

- Significant autoregressive terms at lags 2, 3, and –4  
- COVID-19 Q2 dummy: **large negative effect** (p < 0.001)  
- 2022Q2 shock: **positive and significant** (p < 0.001)  
- Agriculture GDP lags are **not statistically significant**  

This implies **agriculture does not significantly predict total GDP growth** in the short run.

#### (b) Agriculture GDP equation ($\Delta_4 \log(p_1)$)

- Strong own-lag dependence (lag 1, p < 0.01)
- Total GDP lag 1 has mild predictive power (p = 0.036)
- 2022Q2 shock borderline significant (p = 0.099)

This suggests **total GDP growth may influence agricultural GDP**, but the effect is small and short-lived.

### 4.5 Granger Causality Results

Two null hypotheses were tested:

1. **$p_1$ does NOT Granger-cause $p_0$**  
   - p = 0.9539 → fail to reject H0  
   - **Agriculture does not drive total GDP**

2. **$p_0$ does NOT Granger-cause $p_1$**  
   - p = 0.0620 → marginal evidence  
   - **Total GDP may influence agriculture**, but weakly

Instantaneous causality tests showed **no contemporaneous correlation** between the shocks in both equations.

### 4.6 Interpretation of Diagnostics

- System stability: All VAR roots < 1 → model is stable  
- Residual correlation ≈ 0 → equations are properly separated  
- Good fit for agriculture equation (Adjusted $R^2 = 0.73$)  
- Strong improvement when exogenous dummies are included  
  (without dummies, model fit decreases significantly)

### 4.7 Economic Interpretation

- Agriculture GDP does **not** Granger-cause total GDP  
  → Agriculture sector reacts to the overall economy rather than driving it  
- Total GDP shows minor predictive power for agriculture  
  → Agriculture behaves as a dependent subsector  
- COVID-19 had a **severe, short-run negative impact** on total GDP  
- The 2022Q2 shock had a **positive impact**, possibly due to reopening, stimulus, or commodity price recovery

### 4.8 Conclusion: Why VARMAX(4) is Appropriate

The VARMAX(4) model is suitable because:

1. The series are **non-stationary in levels but stationary in seasonal log-differences**  
2. There is **no cointegration**, so VECM is inappropriate  
3. VARselect indicates **lag 4** is optimal  
4. COVID-19 and 2022Q2 shocks create structural breaks → handled via exogenous terms  
5. The model provides  
   - Granger causality  
   - Impulse response functions  
   - Forecast error variance decomposition (FEVD)  
   → allowing quantitative interpretation of agriculture’s contribution to GDP

Overall, the VARMAX(4) framework provides a statistically consistent and economically meaningful way to analyse dynamic interactions between agriculture GDP and total GDP.

```{r}

d4_log_p0 <- diff(log(ts_p0), lag = 4)
d4_log_p1 <- diff(log(ts_p1), lag = 4)

tq <- time(d4_log_p0)  # same time index for both logs
X_var <- cbind(
  covid_q2     = as.numeric(tq == 2020 + 1/4),
  covid_q3     = as.numeric(tq == 2020 + 2/4),
  shock_2022q2 = as.numeric(tq == 2022 + 1/4)
)
p_opt <- 4  # for example
y_var <- cbind(d4_log_p0, d4_log_p1)   # this becomes an "mts" object
colnames(y_var) <- c("d4_log_p0", "d4_log_p1")

cut_num   <- 2023 + 3/4      # 2023Q4 in fractional time
train_ms  <- tq <= cut_num   # logical mask for training
test_ms   <- tq >  cut_num   # logical mask for test

train_y   <- y_var[train_ms, ]
test_y    <- y_var[test_ms, ]

X_train   <- X_var[train_ms, , drop = FALSE]
X_test    <- X_var[test_ms,  , drop = FALSE]

library(vars)

var_train <- VAR(
  y     = train_y,
  p     = p_opt,
  type  = "const",
  exogen = X_train      # not X_var!
)

summary(var_train)

h <- nrow(test_y)

fc_var <- predict(
  var_train,
  n.ahead = h,
  dumvar  = X_test          # exogenous dummies over test horizon
)

# Extract point forecasts
fc_d4_p0 <- fc_var$fcst$d4_log_p0[, "fcst"]
fc_d4_p1 <- fc_var$fcst$d4_log_p1[, "fcst"]

## 2. Build a date index for the test period (2024–2025)
t_test <- tq[test_ms]

q_to_month <- function(q) c(1, 4, 7, 10)[q]
yy <- floor(t_test)
qq <- (round((t_test - yy) * 4) %% 4) + 1
dates_test <- as.Date(sprintf("%04d-%02d-01", yy, q_to_month(qq)))

## 3. Tidy data frame for plotting
df_plot <- tibble(
  date     = rep(dates_test, 2),
  series   = rep(c("d4_log_p0", "d4_log_p1"), each = h),
  actual   = c(test_y[, "d4_log_p0"], test_y[, "d4_log_p1"]),
  forecast = c(fc_d4_p0,             fc_d4_p1)
)

## 4. Plot actual vs forecast (2024–2025)
ggplot(df_plot, aes(x = date)) +
  geom_line(aes(y = actual, colour = "Actual"), linewidth = 1) +
  geom_line(aes(y = forecast, colour = "Forecast"), linewidth = 1, linetype = "dashed") +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(
    title = "VAR forecast vs actual (2024–2025)",
    x = "Date", y = "Quarterly seasonal log-difference",
    colour = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## 5. Optional: error metrics for each series
df_errors <- df_plot |>
  mutate(error = actual - forecast) |>
  group_by(series) |>
  summarise(
    RMSE = sqrt(mean(error^2)),
    MAE  = mean(abs(error)),
    MAPE = mean(abs(error / actual)) * 100
  )

print(df_errors)

```
```{r}
library(vars)
library(tidyverse)
library(ggplot2)

## 1. Build seasonal differences in LEVELS (no logs)
d4_p0 <- diff(ts_p0, lag = 4)   # Δ4 p0
d4_p1 <- diff(ts_p1, lag = 4)   # Δ4 p1

# Combine into a bivariate mts
y_var_d4 <- cbind(d4_p0 = d4_p0,
                  d4_p1 = d4_p1)

# Time index for this differenced series
tq_d4 <- time(d4_p0)

## 2. (Optional) intervention dummies on the differenced series
covid_q2_d4     <- as.numeric(tq_d4 == 2020 + 1/4)
covid_q3_d4     <- as.numeric(tq_d4 == 2020 + 2/4)
shock_2022q2_d4 <- as.numeric(tq_d4 == 2022 + 1/4)

X_d4_mat <- cbind(
  covid_q2     = covid_q2_d4,
  covid_q3     = covid_q3_d4,
  shock_2022q2 = shock_2022q2_d4
)

# Make exogenous dummies a ts so we can use window()
X_d4 <- ts(X_d4_mat,
           start = start(d4_p0),
           frequency = frequency(d4_p0))

## 3. Choose lag order for VAR on differenced levels
lag_sel_d4 <- VARselect(y_var_d4, lag.max = 4, type = "const", exogen = X_d4)
lag_sel_d4$selection
p_opt_d4 <- lag_sel_d4$selection["AIC(n)"]  # or SC(n), HQ(n) if you prefer

## 4. Split into train (up to 2023Q4) and test (2024Q1–2025Q4)
train_y_d4 <- window(y_var_d4, end   = c(2023, 4))
test_y_d4  <- window(y_var_d4, start = c(2024, 1))

X_train_d4 <- window(X_d4, end   = c(2023, 4))
X_test_d4  <- window(X_d4, start = c(2024, 1))

h <- nrow(test_y_d4)

## 5. Fit VAR on training sample (levels, seasonal diff)
var_train_d4 <- VAR(train_y_d4,
                    p     = p_opt_d4,
                    type  = "const",
                    exogen = X_train_d4)

summary(var_train_d4)

## 6. Forecast 2024–2025 with VAR
fc_d4 <- predict(var_train_d4,
                 n.ahead = h,
                 dumvar  = X_test_d4)

# Extract point forecasts for each series
fc_d4_p0 <- fc_d4$fcst$d4_p0[, "fcst"]
fc_d4_p1 <- fc_d4$fcst$d4_p1[, "fcst"]

## 7. Build dates for test horizon (2024–2025)
t_test_d4 <- time(test_y_d4)

q_to_month <- function(q) c(1, 4, 7, 10)[q]
yy <- floor(t_test_d4)
qq <- (round((t_test_d4 - yy) * 4) %% 4) + 1
dates_test_d4 <- as.Date(sprintf("%04d-%02d-01", yy, q_to_month(qq)))

## 8. Tidy data for plotting actual vs forecast
df_plot_d4 <- tibble(
  date     = rep(dates_test_d4, 2),
  series   = rep(c("d4_p0 (Δ4 GDP)", "d4_p1 (Δ4 Agri)"), each = h),
  actual   = c(test_y_d4[, "d4_p0"], test_y_d4[, "d4_p1"]),
  forecast = c(fc_d4_p0,            fc_d4_p1)
)

## 9. Plot: actual vs forecast for 2024–2025
ggplot(df_plot_d4, aes(x = date)) +
  geom_line(aes(y = actual, colour = "Actual"), linewidth = 1) +
  geom_line(aes(y = forecast, colour = "Forecast"), linewidth = 1,
            linetype = "dashed") +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(
    title = "VAR on Δ4 levels: forecast vs actual (2024–2025)",
    x = "Date",
    y = "Seasonal difference (Δ4 in levels)",
    colour = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## 10. Error metrics (RMSE, MAE, MAPE) for each series
df_errors_d4 <- df_plot_d4 |>
  mutate(error = actual - forecast) |>
  group_by(series) |>
  summarise(
    RMSE = sqrt(mean(error^2)),
    MAE  = mean(abs(error)),
    MAPE = mean(abs(error / actual)) * 100
  )

print(df_errors_d4)

```

```{r}

```

